---
title: "An Introduction to Rcpp"
subtitle: "Stat 696: Machine Learning"
author: "Andrew Bates"
date: "02/14/2019"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# What Is Rcpp?

--

An R package connecting R and C++

--

You write C++ code

--

Rcpp handles the details


---

# Why Would I Use It?

--

You have some R code but...

--

It's too slow

--

You've tried vectorizing, etc.

--

It's still too slow

--

Some use cases (from [Advanced R](http://adv-r.had.co.nz/Rcpp.html))

- Loops that can't be vectorized (e.g. MCMC)
- Need to call R functions millions of times (e.g. recursive functions)
- Need complex data structures not available in R

---

# A First Example

Write some C++ code

--

```{Rcpp, cache = TRUE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector square_root(NumericVector x) {
  return sqrt(x);
}
```

--

Then call it from R

--

```{r }
x <- 1:5
square_root(x)
```


---

# Let's Break It Down

```{Rcpp, eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector square_root(NumericVector x) {
  return sqrt(x);
}
```

--

.pull-left[

- `#include <Rcpp.h>`

- `using namespace Rcpp;`

- `// [[Rcpp::export]]`
 
- `NumericVector square_root(...`

]

.pull-right[

- like `library(mypackage)`

- so we don't have to write `package::function`

- send function to R

- function definition

]

???
things to note:
1. semicolons
2. need to specify types

---

# A More Realistic Example

Draw sample from multivariate normal

```{Rcpp, cache = TRUE}
#include <RcppEigen.h>
// [[Rcpp::depends(RcppEigen)]]

using namespace Rcpp;
using namespace Eigen;

// [[Rcpp::export]]
VectorXd mv_normal(VectorXd mu, MatrixXd sigma){
  int r = mu.size();
  VectorXd standard_normal(r);
  VectorXd result(r);

  standard_normal = as<VectorXd>(rnorm(r, 0, 1));
  result = mu + sigma.llt().matrixL() * standard_normal;
  return result;
}
```


???
things to note:
1. using RcppEigen - wrapper for Eigen C++ library for linear algebra
2. `rnorm` is part of Rcpp sugar - C++ functions that will be familiar to R users
3. using Cholesky decomposition

---

# A More Realistic Example

In R:

```{r}
mu <- c(4,4)
sigma <- diag(2) * 2
mv_normal(mu, sigma)
```


---

class: middle

Let's see this in action!


---

# Resources

[Rcpp.org](http://www.rcpp.org)

[Rcpp Gallery](http://gallery.rcpp.org) - *lots* of examples

Rcpp chapter of [Advanced R](http://adv-r.had.co.nz/Rcpp.html)

Compiled Code chapter of [R Packages](http://r-pkgs.had.co.nz/src.html)



---

class: center, middle

# Happy Rcpping!

These slides were created with the R package [**xaringan**](https://github.com/yihui/xaringan).

You can find them at [https://asbates.rbind.io/slides/rcpp-intro](https://asbates.rbind.io/slides/rcpp-intro)

You can find code for the examples at [https://github.com/asbates/neural-nets/tree/master/rcpp-intro](https://github.com/asbates/neural-nets/tree/master/rcpp-intro) 



